---
title: "LOG708 - Chapter 3 Exercises (C3.1 & C3.2)"
author: "Alessandro Di Mattia"
date: "`r Sys.Date()`"
output: html_document
---

Due: Thursday 03 Sep 2026 (Week 36 lab)
Source: [Log708 Compendium — Exercises, Introduction to R](https://home.himolde.no/arntzen/Log708-compendium/exercises/_book/introduction-to-r.html)

## Exercise 3.1 — Installation and Setup

> **Note:** If you are working on a campus PC, do **not** try to install R/RStudio there.

- [ ] Install R (from CRAN) and RStudio. Watch the RStudio installation video on Canvas.
- [ ] Watch "More about RStudio" and set up an organized folder structure for coursework
      (see the sample `MASTER_Coursework` folder on Canvas).
- [ ] Identify a secure backup location for your coursework (protect against hardware
      failure or theft — e.g. cloud storage, external drive).
- [ ] Watch "RStudio projects" and create an R Project for the Chapter 3 exercises,
      using the folder structure set up above.
- [ ] (Optional) Set a default working directory via `Tools → Global Options`.

## Exercise 3.2 — Variables and Vectors

- [ ] Watch the "Variables and Vectors" video (Canvas/Panopto) and download the
      accompanying R script.
- [ ] Test the code yourself in your own RStudio installation.
- [ ] Learn the keyboard shortcuts: `Ctrl+Enter` (run line/selection), `Ctrl+C`, etc.

### 1. Birth year

```{r}
birth_year <- 2000
birth_year
```

### 2. Date vector

```{r}
dates <- as.Date(c("2026-01-01", "2026-06-15", "2026-12-31"))
dates
```

### 3. Basic calculations

```{r}
current_year <- 2026
age <- current_year - birth_year
age
```

### 4. Logical operators

```{r}
age > 18
birth_year == 2000
```

### 5. Create a vector

```{r}
v <- c(10, 20, 30, 40, 50) 
v
```

### 6. Indexing

```{r}
v[1]      # first element
v[2:4]    # elements 2 to 4
v[-1]     # all but the first
```

### 7. Manipulating vector data

```{r}
v[6] <- 60      # add an element
v <- v[v != 30] # remove an element
v
```

## Exercise 3.4

```{r}

DF <- data.frame(name = c("Norway", "Sweden", "Denmark", "Finland", 
                          "Germany", "Holland", "Switzerland", "Poland"),
                 population = c(5.5, 10.7, 5.9, 5.5, 84.4, 17.6, 8.8, 40),
                 GDP = c(593, 593, 400, 300, 4456, 991, 885, 1801))
DF

names(DF)[3] <- "gdp"
# names(DF)[names(DF) == "GDP"] <-"gdp"
# install.packages("dplyr")
library(dplyr)
# DF <- DF |> rename(gdp = GDP)


```

