---
title: "LOG708 - Chapter 4 Exercises (4.1 - 4.5)"
author: "Alessandro Di Mattia"
date: "`r Sys.Date()`"
output: html_document
---

Source: [Log708 Compendium — Exercises, Hypothesis Testing](https://home.himolde.no/arntzen/Log708-compendium/exercises/_book/hypothesis-testing.html)

In all tests, use significance level $\alpha = 0.05$ unless otherwise specified.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4.1 — Testing the Mean and a Proportion

In a supermarket, the manager tries to forecast and plan operations for the next
quarter. One assumption is that the mean spending $\mu$ per customer is 290 NOK. A
second assumption is that only 20% of customers pay with cash, while 10% of
customers make a cash withdrawal using a bank card. In a sample of 400 customers,
the average spending was found to be $\bar{x} = 319$ NOK, with a standard deviation
of $S_x = 100$. Also in the sample, 56 of the 400 customers payed with cash.

### a. Test H0: mu = 290 vs H1: mu != 290

Using $\mu = 290$ as above, and $N(0,1)$ as the approximate null distribution for
this one-sample t-test:

```{r}
n <- 400
x_bar <- 319
s_x <- 100

t_obs <- (x_bar - 290) / (s_x / sqrt(n))
t_obs

# two-sided p-value
pval <- 2 * pnorm(-abs(t_obs))
pval
```

The observed value is 5.8, and the P-value is practically 0, so we reject $H_0$ at
the 0.05 level. Mean spending is very unlikely to be 290 NOK.

### b. Test H0: p = 0.20 vs H1: p != 0.20

Let $p$ be the proportion of customers paying cash.

```{r}
x <- 56       # number paying cash
n <- 400

phat <- x / n
phat

z_obs <- (phat - 0.20) / sqrt(0.20 * (1 - 0.20) / n)
z_obs

pval_p <- 2 * pnorm(-abs(z_obs))
pval_p
```

We get $Z_{obs} = -3.0$ and P-value $\approx 0.003$, so we again reject $H_0$. The
proportion of cash-paying customers is likely quite a bit lower than the assumed
0.20. We can double check this with R's built-in test:

```{r}
prop.test(x, n, p = 0.20, correct = FALSE)
```

### c. Is the cash balance reasonable?

Suppose 10% of customers make a cash withdrawal, averaging 400 NOK, while (by
assumption) 20% pay with cash, averaging 319 NOK. For each withdrawal (-400), there
are on average two cash payments (2 * 319 = 638 NOK), so the cash balance should be
fine under the original assumptions. Even using the estimated 14% paying cash
(instead of the assumed 20%), each -400 withdrawal is countered by
1.4 * 319 = 447 NOK in cash payments, so the balance still looks reasonable.

## Exercise 4.2 — One-Sample Tests, Confidence Intervals and Correlation

Data: `Trip_durations.csv`. Let $\mu_d$ be the (true) mean duration of trips, and
$\mu_s$ be the mean distance of trips.

```{r}
tripdata <- read.csv("Data/Trip_durations.csv")
head(tripdata)
```

### a. Sample size, mean and standard deviation for Duration

```{r}
n <- nrow(tripdata)
x_bar <- mean(tripdata$Duration)
s_x <- sd(tripdata$Duration)

# print the three numbers as a vector (just to save space in output)
c(n, x_bar, s_x)
```

### b. Test H0: mu_d = 24 vs H1: mu_d != 24 (without t.test)

The test statistic is
$$ T = \frac{\bar{x} - 24}{S_x / \sqrt{n}} $$
with $N(0,1)$ as the approximate null distribution.

```{r}
t_obs <- (x_bar - 24) / (s_x / sqrt(n))
t_obs

p_value <- 2 * pnorm(-abs(t_obs))
p_value
```

The P-value is about 0.52, so with $\alpha = 0.05$ we can **not** reject $H_0$.

### c. Confirm with t.test

```{r}
durtest1 <- t.test(tripdata$Duration, mu = 24, alternative = "two.sided")
durtest1$statistic
durtest1$p.value
```

We get almost identical values to the manual calculation in b.

### d. Repeat b. and c. for H0: mu_d = 23

```{r}
t_obs <- (x_bar - 23) / (s_x / sqrt(n))
t_obs

p_value <- 2 * pnorm(-abs(t_obs))
p_value

durtest2 <- t.test(tripdata$Duration, mu = 23, alternative = "two.sided")
durtest2$statistic
durtest2$p.value
```

This time the P-value is about 0.007, well below 0.05, so we clearly reject $H_0$ —
the opposite conclusion from testing $\mu_d = 24$.

### e. Test whether mu_s > 10 (Distance)

```{r}
disttest <- t.test(tripdata$Distance, mu = 10, alternative = "greater")
disttest
```

We get $T_{obs} \approx 1.95$ with P-value $\approx 0.026$, which is below 0.05, so
we reject $H_0$ and conclude the mean distance is greater than 10 km — although the
evidence is relatively weak (the P-value is not much below 0.05).

### f. 95% confidence intervals for mu_d and mu_s

```{r}
dur_test <- t.test(tripdata$Duration)
dist_test <- t.test(tripdata$Distance)

dur_test$conf.int
dist_test$conf.int
```

The duration CI contains 24 but not 23, matching the "not rejected" / "rejected"
results in b/c and d respectively. The distance CI reaches from about 10.00 to
10.81, consistent with the (weak) rejection in e.

### g. Correlation between Distance and Duration

```{r}
with(tripdata, cor(Duration, Distance))

with(tripdata, plot(Distance, Duration,
                    main = "Duration vs Distance for road trips."))
```

The correlation is about 0.95, and the scatterplot shows a strong, clearly positive
relationship between the two variables.

## Exercise 4.3 — Comparing Groups (Two-Sample t-tests)

Data: `flat_prices.csv`, with sold flats in Molde, Kristiansund and Ålesund. The
`town` variable is encoded 1 = Molde, 2 = Kristiansund, 3 = Ålesund.

```{r}
flats <- read.csv("Data/flat_prices.csv")
head(flats)
```

### a. Mean price in the three towns

```{r}
with(flats, tapply(price, town, mean))
```

We see clear differences in average price between the towns.

### b. Boxplot of price by town

```{r}
with(flats, boxplot(price ~ town))
```

This shows the same pattern of differences, and also a somewhat wider spread of
prices in Ålesund.

### c. Two-sample t-tests: Molde vs Kristiansund, and Kristiansund vs Ålesund

$H_0: \mu_M = \mu_K$ vs $H_1: \mu_M \neq \mu_K$. Since `t.test` needs a grouping
variable with only two levels, we first exclude town 3.

```{r}
flats_MK <- subset(flats, town != 3)
with(flats_MK, t.test(price ~ town))
```

The relatively high P-value (about 0.26) means we can not reject $H_0$: no
significant nominal price difference between Molde and Kristiansund.

```{r}
flats_KA <- subset(flats, town != 1)
with(flats_KA, t.test(price ~ town))
```

The P-value here (about 0.052) is right at the boundary — close to significant, but
still not below 0.05, so we do not reject $H_0$.

### d. Mean square meter size in the three towns

```{r}
with(flats, tapply(area, town, mean))
```

The sample flats are, on average, somewhat different in size across towns, with
Kristiansund flats being the largest on average.

### e. Correlation between area and price

```{r}
with(flats, cor(price, area))

with(flats, plot(area, price))
```

The correlation is strong, about 0.95, and the scatterplot confirms a clear
positive relationship, as expected.

### f. New variable: square meter price

```{r}
flats$sqmprice <- flats$price / flats$area
head(flats)
```

### g. Two-sample t-test on square meter price: Molde vs Kristiansund

```{r}
flats_MK <- subset(flats, town != 3)
with(flats_MK, t.test(sqmprice ~ town))
```

The P-value is almost 0, far below 0.05, so this time we **reject** $H_0$: square
meter prices are significantly higher in Molde than in Kristiansund.

### h. Why do nominal and square-meter price comparisons differ?

When comparing nominal prices, we ignore the fact that, on average, Molde flats are
also smaller than Kristiansund flats. Nominal price differences can therefore be
misleading, since they mix together a "size effect" and a genuine "price level"
effect. Square meter price controls for size, giving a fairer comparison of the
underlying price level between towns.

### i. Square meter price: Kristiansund vs Ålesund, and Molde vs Ålesund

```{r}
flats_KA <- subset(flats, town != 1)
flats_MA <- subset(flats, town != 2)

with(flats_KA, t.test(sqmprice ~ town))
with(flats_MA, t.test(sqmprice ~ town))
```

We reject $H_0$ when comparing Kristiansund and Ålesund (very low P-value), but not
when comparing Molde and Ålesund (P-value well above 0.05).

### Bonus: using a factor for town

Converting `town` to a factor with meaningful labels makes the output easier to
read:

```{r}
flats$town <- factor(flats$town, levels = 1:3,
                     labels = c("Molde", "Krsund", "Alesund"))
with(flats, tapply(price, town, mean))
```

## Exercise 4.4 — A Binomial Test

Using `flat_prices.csv` again, let $p$ be the (true) proportion of flats sold that
have 2 rooms or less. We test

$$H_0: p = 0.50 \ \text{ vs } \ H_1: p < 0.50 .$$

```{r}
flats <- read.csv("Data/flat_prices.csv")

# TRUE/FALSE indicator for "small" flats: fewer than 3 rooms (i.e. <= 2 rooms)
flats$small <- flats$rooms < 3

countsm <- table(flats$small)
countsm
```

`countsm` shows the count of small (`rooms < 3`) and large (all others) flats — 59
small and 91 large flats.

```{r}
binom.test(countsm["TRUE"], sum(countsm), p = 0.50, alternative = "less")
```

The P-value is far below 0.05, so we reject $H_0$: the true proportion of small
flats is significantly lower than 50%.

## Exercise 4.5 — Checking Normality

Using `flat_prices.csv`, we check whether the selling price of flats appears to
follow a normal distribution.

### a. Would you expect prices to be normally distributed?

Prices in a real estate market are typically **not** exactly normal: they are
bounded below by 0 and tend to have a longer right tail, since a few large/luxury
properties can sell far above the typical price, while very few (if any) sell for
much less than the bulk of the market. This kind of right-skew conflicts with the
symmetric, unbounded shape of the normal distribution.

### b. Normality test for price

```{r}
shapiro.test(flats$price)

qqnorm(flats$price)
qqline(flats$price)
```

A small `shapiro.test` P-value (below 0.05) indicates a significant departure from
normality, and the Q-Q plot typically shows points curving away from the reference
line in the tails — consistent with the right-skew expected in a).

### c. Repeat within each town

```{r}
with(flats, tapply(price, town, shapiro.test))

with(flats, tapply(price, town, function(x) { qqnorm(x); qqline(x) }))
```

The combined sample mixes three sub-populations with different mean price levels
(see Exercise 4.3a), so pooling them together can make the overall distribution
look more skewed/irregular than it does within any single town. Testing normality
separately per town removes this "mixture" effect, and each town's own price
distribution is closer to (though still not perfectly) normal.

### d. Are square meter prices normally distributed?

```{r}
flats$sqmprice <- flats$price / flats$area

shapiro.test(flats$sqmprice)

qqnorm(flats$sqmprice)
qqline(flats$sqmprice)
```

Since `sqmprice` already controls for flat size (and implicitly removes some of the
town-to-town price-level differences relative to nominal price), it tends to be
closer to normally distributed than the raw `price` variable — though it should
still be checked against the `shapiro.test` P-value and Q-Q plot rather than
assumed.
