---
title: "LOG708 - Chapter 3 Exercises (3.1 - 3.10)"
author: "Alessandro Di Mattia"
date: "`r Sys.Date()`"
output: html_document
---

Due: Thursday 03 Sep 2026 (Week 36 lab)
Source: [Log708 Compendium — Exercises, Introduction to R](https://home.himolde.no/arntzen/Log708-compendium/exercises/_book/introduction-to-r.html)

## Setup — Install and load required packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
# packages used throughout this document
required_packages <- c("dplyr", "ggplot2")

# install any that are missing, then load them all
missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

## Exercise 3.1 — Installation and Setup

> **Note:** If you are working on a campus PC, do **not** try to install R/RStudio there.

- [ ] Install R (from CRAN) and RStudio. Watch the RStudio installation video on Canvas.
- [ ] Watch "More about RStudio" and set up an organized folder structure for coursework
      (see the sample `MASTER_Coursework` folder on Canvas).
- [ ] Identify a secure backup location for your coursework (protect against hardware
      failure or theft — e.g. cloud storage, external drive).
- [ ] Watch "RStudio projects" and create an R Project for the Chapter 3 exercises,
      using the folder structure set up above.
- [ ] (Optional) Set a default working directory via `Tools → Global Options`.

## Exercise 3.2 — Variables and Vectors

- [ ] Watch the "Variables and Vectors" video (Canvas/Panopto) and download the
      accompanying R script.
- [ ] Test the code yourself in your own RStudio installation.
- [ ] Learn the keyboard shortcuts: `Ctrl+Enter` (run line/selection), `Ctrl+C`, etc.

### 1. Birth year

```{r}
birth_year <- 2000
birth_year
```

### 2. Date vector

```{r}
dates <- as.Date(c("2026-01-01", "2026-06-15", "2026-12-31"))
dates
```

### 3. Basic calculations

```{r}
current_year <- 2026
age <- current_year - birth_year
age
```

### 4. Logical operators

```{r}
age > 18
birth_year == 2000
```

### 5. Create a vector

```{r}
v <- c(10, 20, 30, 40, 50)
v
```

### 6. Indexing

```{r}
v[1]      # first element
v[2:4]    # elements 2 to 4
v[-1]     # all but the first
```

### 7. Manipulating vector data

```{r}
v[6] <- 60      # add an element
v <- v[v != 30] # remove an element
v
```

### 8. Assign 1, 2, 3 to a, b, c

```{r}
a <- 1
b <- 2
c <- 3
```

### 9. Test whether a is equal to b

```{r}
a == b
```

### 10. Test whether a+b = c

```{r}
a + b == c
```

### 11. Vector f

```{r}
f <- c(12, 13, 14, 15)
f
```

### 12. Vector y, the product of b and each element of f

```{r}
y <- b * f
y
```

### 13. Vectors u and v

```{r}
u <- c(55, 29, 51, 35, 33, 42)
v <- c(2, 4, 6, 8, 10, 12)
```

### 14. 4th element of u assigned to x

```{r}
x <- u[4]
x
```

### 15. Change the first element of u to 25

```{r}
u[1] <- 25
u
```

### 16. Logical vector z: u < 35

```{r}
z <- (u < 35)
z
```

### 17. Logical vector w: u < 35 and v > 5

```{r}
w <- (u < 35) & (v > 5)
w
```

### 18. Length and type of u and w

```{r}
length(u)
str(u)
length(w)
str(w)
```

### 19. Vectorized operations

```{r}
u + v
u * v
u / v
v^4
```

### 20. List the global environment

```{r}
ls()
```

## Exercise 3.3 — Functions and Sequences

```{r}
# make a vector z of 5-10 arbitrary numbers
z <- c(5, 67, 33, 43, 41, 55)   # or try z <- sample(1:100, 10)
z

# sum of elements in z
sum(z)

# average value of z
mean(z)

# median value of z
median(z)

# mean value of z^2
mean(z^2)
```

```{r}
# factorial(10) = 10*9*8*...*3*2*1
factorial(10)

# factorial(100)
factorial(100)
```

```{r}
# Suppose X is a standard normal variable. Find the probability that X < 2.5
pnorm(2.5)

# Find the probability that X > 2.5 (law of complement, or lower.tail = FALSE)
1 - pnorm(2.5)
pnorm(2.5, lower.tail = FALSE)
```

```{r}
# Suppose W is normal with mean 4000, sd 300. Find P(W < 4500) and P(W > 5000)
pnorm(4500, mean = 4000, sd = 300)
pnorm(5000, mean = 4000, sd = 300, lower.tail = FALSE)

# Find all probabilities that W < x, for x = 4100, 4200, 4300, 4400, 4500 (vectorized)
x <- c(4100, 4200, 4300, 4400, 4500)
pnorm(x, mean = 4000, sd = 300)
```

```{r}
# write R code to generate the following sequences

# 1,3,5,7,9,11
seq(from = 1, to = 11, by = 2)

# 5,6,7,8,9,10
5:10

# 1,3,1,3,1,3,1,3,1,3
rep(c(1, 3), 5)   # or rep(c(1, 3), times = 5)

# 8,7,6,5,4,3,2,1
8:1
```

```{r}
# suppose y is the vector defined by
y <- c(5, 6, 12, 53, 15, 95, 12, 51, 25, 34, 59, 15, 40, 34)

# find the length of y
length(y)

# extract a vector with the last 6 elements of y
y[9:14]
```

## Exercise 3.4 — Data Frames: Basics

```{r}
DF <- data.frame(name = c("Norway", "Sweden", "Denmark", "Finland",
                          "Germany", "Holland", "Switzerland", "Poland"),
                 population = c(5.5, 10.7, 5.9, 5.5, 84.4, 17.6, 8.8, 40),
                 GDP = c(593, 593, 400, 300, 4456, 991, 885, 1801))

# view the dataframe in the console (and in a spreadsheet-like view with View(DF))
DF
```

```{r}
# use R functions to find number of rows, number of columns
c(nrow(DF), ncol(DF))

# get a vector with the names of variables in DF
names(DF)

# play with head and tail to show first and last 3 lines of DF
head(DF, 3)
tail(DF, 3)
```

```{r}
# rename column "GDP" to "gdp"
names(DF)[3] <- "gdp"
names(DF)
```

```{r}
# find the median GDP and the mean population in DF
median(DF$gdp)
mean(DF$population)
```

```{r}
# extract the column "population" into a separate vector "pop" (outside of DF)
pop <- DF$population

# extract subset A of DF with countries of population greater than 8 million
A <- subset(DF, population > 8)
A

# extract subset B of DF where population is greater than 8 and gdp < 1000
B <- subset(DF, population > 8 & gdp < 1000)
B

# how can you also extract subset B from A?
B2 <- subset(A, gdp < 1000)
B2
```

```{r}
# extract row 5 from DF
DF[5, ]

# extract a dataframe with rows 2,3,4 from DF
DF[2:4, ]

# extract a dataframe with columns 2, 3 from DF
DF[, 2:3]

# extract a dataframe with columns 1 and 3 from DF
DF[, c(1, 3)]

# extract a dataframe with rows 1-4 and columns 1 and 3 from DF
DF[1:4, c(1, 3)]
```

```{r}
# run this code, but try first to guess what it does
subset(DF, startsWith(name, "S"))
```

```{r}
# compute a new column gdppc, containing GDP per capita for the countries
DF$gdppc <- DF$gdp / DF$population
DF
```

Population is in millions and GDP is in billions USD, so `gdppc` is in units of
1000 USD. For example Denmark has about 68 000 USD in GDP per capita.

```{r}
# use "summary" function to summarize the data in DF
summary(DF)
```

### b. Sorting data frames

```{r}
# example from the exercise text: sort DF by population
S <- order(DF$population)
DF[S, ]
```

```{r}
# sort DF by gdp
Sg <- order(DF$gdp)
DF[Sg, ]

# sort DF by gdp, decreasing
Ug <- order(DF$gdp, decreasing = TRUE)
DF[Ug, ]

# sort DF by name
Vn <- order(DF$name)
DF[Vn, ]
```

Sorting by `name` sorts the rows alphabetically, since `order()` works on character
vectors just as well as on numeric ones.

## Exercise 3.5 — Visualization

### a. Scatterplot of population vs gdp

```{r}
with(DF, plot(population, gdp, main = "Population and GDP",
              xlab = "Population",
              ylab = "GDP"))
```

### b. Save the plot as a .pdf file

```{r, eval=FALSE}
pdf("population_vs_gdp.pdf")
with(DF, plot(population, gdp, main = "Population and GDP",
              xlab = "Population",
              ylab = "GDP"))
dev.off()
```

### c. Boxplot of gdppc

```{r}
with(DF, boxplot(gdppc, main = "Boxplot of GDP per capita."))
```

### d. Histogram of gdppc

```{r}
with(DF, hist(gdppc, main = "Histogram of GDP per capita."))
```

### e. (Optional) ggplot2 version

```{r, eval=FALSE}
# install.packages("ggplot2")
library(ggplot2)
ggplot(DF, aes(x = population, y = gdp)) +
  geom_point() +
  labs(title = "Population and GDP", x = "Population", y = "GDP")
```

## Exercise 3.6 — NA's and Factors

### a. Set Poland's GDP to NA

```{r}
DF[8, "gdp"] <- NA
DF
```

### b. Deal with computations when NA's are present

```{r}
mean(DF$gdp, na.rm = TRUE)
```

### c. Run summary on DF

```{r}
summary(DF)
```

`summary()` reports the number of `NA`'s for each variable that contains missing
values, in addition to the usual statistics computed on the non-missing values.

### d. Table and barplot of 20 coin/die tosses

```{r}
toss <- c(4,4,3,5,3,6,3,5,4,3,4,3,4,1,3,1,5,5,4,3,4,1,6)

t <- table(toss)
t
barplot(t)
```

The value "2" never appears in the observed tosses, so `table()` simply does not
report a bar for it — a viewer might wrongly conclude that "2" is not a possible
outcome. This can be fixed by first converting `toss` to a `factor` with all
possible levels declared:

```{r}
tossfac <- factor(toss, levels = 1:6)
t <- table(tossfac)
t
barplot(t)
```

### e. Probability of no 2's in 20 tosses

```{r}
prob <- (5/6)^20
prob
```

Assuming a fair die and independent tosses, the probability of *not* getting a "2"
on a single toss is 5/6. For this to happen 20 times in a row we multiply the
probability for each toss, giving about 0.026 — i.e. it happens roughly 1 out of
every 40 such sequences.

## Exercise 3.7 — Descriptive Statistics

```{r}
# find information about the data, let the name AQ refer to the dataframe
?airquality
AQ <- airquality
```

### a. Basic info about the data frame

```{r}
nrow(AQ)
ncol(AQ)
names(AQ)
head(AQ)
tail(AQ)
```

### b. Rename "Solar.R" to "SolarRad"

```{r}
names(AQ)[2] <- "SolarRad"
names(AQ)
```

### c. Summary of AQ

```{r}
summary(AQ)
```

There are 37 `NA`'s in `Ozone`, and 7 `NA`'s in `SolarRad`.

### d. Mean temperature by month

```{r}
mtemp <- with(AQ, tapply(Temp, Month, mean))
mtemp
barplot(mtemp, main = "Mean temperature by month.")
```

We see the highest mean temperatures in July and August.

### e. Daily temperature time series

```{r}
with(AQ, plot(Temp,
              type = "l",
              xlab = "Day",
              main = "Daily temperature (F)",
              lwd = 2))
```

Since one temperature observation is plotted per day (with time along the first
axis), we get what is usually called a "time series plot".

### f. Mean solar radiation by month (with NA's)

```{r}
with(AQ, tapply(SolarRad, Month, mean))

# avoid NA results with na.rm = TRUE
with(AQ, tapply(SolarRad, Month, mean, na.rm = TRUE))
```

### g. Barplot of mean solar radiation by month

```{r}
solradmean <- with(AQ, tapply(SolarRad, Month, mean, na.rm = TRUE))
barplot(solradmean, main = "Mean solar radiation by month",
        xlab = "Month",
        ylab = "Radiation")
```

### h. Convert temperature to Celsius

```{r}
AQ$celsiustemp <- (5/9) * (AQ$Temp - 32)
```

### i. Boxplots of Celsius temperature

```{r}
# all data in one boxplot
with(AQ, boxplot(celsiustemp, main = "Boxplot of temperature"))

# grouped by month
with(AQ, boxplot(celsiustemp ~ Month, main = "Boxplot of temperature, by month"))
```

### j. Correlation between wind and temperature

```{r}
with(AQ, cor(Wind, Temp))
with(AQ, cor(Wind, celsiustemp))

with(AQ, plot(Wind, celsiustemp, main = "Temperature vs wind",
              ylab = "Temperature"))
```

It does not matter whether temperature is measured in Fahrenheit or Celsius degrees,
since correlation is invariant to a linear (affine) change of scale.

```{r}
# get the 6 first columns (the original, un-derived variables)
AQorig <- AQ[, 1:6]

# get the correlation matrix, ignoring NA observations
cor(AQorig, use = "complete.obs")
```

## Exercise 3.8 — Working Directory, Reading and Writing Data Files

### a. Working directory

```{r}
getwd()
```

### b. Download and extract log708data

- [x] Download `log708data.zip` (or `.7z`) from Canvas and extract it into the
      `Data` folder in the LOG708 folder structure.

### c. Read flights_NO.csv

```{r}
flights_NO <- read.csv("Data/flights_NO.csv")
head(flights_NO)
```

### d. Subset of Monday flights with more than 50 seats

```{r}
df2 <- subset(flights_NO, Day.of.Week == "Monday" & Seats > 50)

nrow(flights_NO)
nrow(df2)
```

`df2` contains the flights departing on Mondays, using airplanes with more than 50
seats.

### e. Write df2 to disk

```{r}
# write df2 to disk with a given file name
write.csv(df2, "LargeMondayFlights.csv", row.names = FALSE)

# view the working directory, the new file should be there
dir()
```

### f. Read the file back in and compare

```{r}
df3 <- read.csv("LargeMondayFlights.csv")
identical(df2, df3)

# to make them truly identical, keep row names when writing/reading
write.csv(df2, "LargeMondayFlights.csv")
df3 <- read.csv("LargeMondayFlights.csv", row.names = 1)
identical(df2, df3)
```

### g. Write df2 to the main data folder

```{r}
write.csv(df2, "Data/LargeMondayFlights_NO.csv")
```

## Exercise 3.9 — R Packages

```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
# already installed and loaded in the "Setup" chunk above, but this is the
# command you need to run once per R session to activate the package:
library(ggplot2)

# include "mtcars" as a dataframe in the workspace
data(mtcars)
head(mtcars)
```

```{r}
# Redefine variable "am" as a factor, just to make figure labels look right.
shift <- factor(mtcars$am, labels = c("automatic", "manual"))

# make a plot
ggplot(mtcars, aes(x = wt, y = hp, color = shift)) +
  geom_point(size = 3) +
  labs(title = "Horsepower vs weight",
       x = "Weight",
       y = "Horsepower")
```

### a. Relationship between weight and horsepower

Horsepower is positively correlated with weight, as expected. Automatic-shift cars
are mainly the heavier vehicles in this data set.

### b. Add a regression line per group

```{r}
ggplot(mtcars, aes(x = wt, y = hp, color = shift)) +
  geom_point(size = 3) +
  labs(title = "Horsepower vs weight",
       x = "Weight",
       y = "Horsepower") +
  geom_smooth(method = lm, se = FALSE)
```

`geom_smooth(method = lm, se = FALSE)` adds a fitted linear regression line for
each group defined by the `shift` variable.

## Exercise 3.10 — Getting Help

Suppose we have a standard normal variable Z, and another normal variable X with
mean = 10 and standard deviation = 3. We want to find
P[Z ≤ 2.1], P[Z ≥ 2.1], P[X ≤ 12], P[8.4 ≤ X ≤ 10.9].

### a. Using ?Normal / help(Normal)

```{r}
pnorm(2.1)
1 - pnorm(2.1)
pnorm(2.1, lower.tail = FALSE)

pnorm(12, mean = 10, sd = 3)

pnorm(10.9, mean = 10, sd = 3) - pnorm(8.4, mean = 10, sd = 3)
```

### b. Searching the web

- [ ] Search for "R calculate normal distribution" (or ask an AI assistant) and
      compare the results to part a.

### c. Find b such that P[Z ≤ b] = 0.80

```{r}
qnorm(0.80)
```

### d. Find c such that P[X ≥ c] = 0.90

```{r}
qnorm(0.10, mean = 10, sd = 3)

# equivalently, using lower.tail = FALSE
qnorm(0.90, mean = 10, sd = 3, lower.tail = FALSE)
```

### e. Simulate a sample from a normal distribution

```{r}
set.seed(3333)
my_x <- rnorm(30, mean = 10, sd = 3)

# compute mean and standard deviation
M <- mean(my_x)
S <- sd(my_x)

# show result
c(M, S)
```

The sample mean and standard deviation are reasonably close to the "true"
parameter values 10 and 3.
