---
title: "assignment"
author: "Phinehas Yitschaq Benneh"
date: "2026-10-01"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("C:/Users/Phinehas/Desktop/log 708 assignment 1")
```

```{r}
cardata <- read.csv("el_car_data_1.csv")
```
# task a

```{r}
head(cardata,10)
tail(cardata, 10)
nrow(cardata)
```

# task b
```{r}
cardata$diesel_pct <- (cardata$diesel/cardata$total)*100
```


```{r}
cardata$gasoline_pct <- (cardata$gasoline/cardata$total)*100
```

# task c
```{r}
cardata$zone = ordered(cardata$zone, levels = c("city",
"below30",
"30to60",
"above60"))
```
# task c1

```{r}
head(cardata$zone)
```

# task d
```{r}
cardata12 <- subset(cardata, year==2012)
cardata18 <- subset(cardata, year==2018)
```

# task e
```{r}
table(cardata18$zone)
```
```{r}
table(cardata18$zone)
barplot(table(cardata18$zone))
```



# task f
```{r}
sorted_data <-cardata18[order(cardata18$total),]
head(sorted_data, 1)
tail(sorted_data,1)
```
# task g 
```{r}
with(cardata, tapply(income_med_adj,year, mean, na.rm=TRUE))
```
# task h
```{r}
S <- order(cardata12$pct_elcar, decreasing = TRUE)
```
# look at the first 10
```{r}
S[1:10]
```
#extract the rows in cardata12 corresponding to the first 10 indexes.
```{r}
cardata12[S[1:10],]
```
# task i
```{r}
S_18 <- order(cardata18$pct_elcar, decreasing = TRUE)
```
# task j
```{r}
mean_by_zone <- aggregate(pct_elcar~zone, data = cardata, FUN = mean, na.rm = TRUE)
```
# task k
```{r}
boxplot(pct_elcar~ zone, data= cardata, main="Distribution of Electric Car % by zone", xlab= "zone", ylab= "Electric Car Percentage(pct_elcar)",col= "lightblue", las= 1)
```
#task l, calculate total electric cars and total cars in Norway
```{r}
total_el_2012 <- sum(cardata$el, na.rm = TRUE)
toatl_cars_2012 <- sum(cardata12$total, na.rm = TRUE)
```
# create national percentage for 2012
```{r}
national_pct_2012 <- (total_el_2012/ toatl_cars_2012)* 100
```
# calculate total electric cars and total cars in norway
```{r}
total_el_2018 <- sum(cardata18$el, na.rm = TRUE)
total_cars_2018 <- sum(cardata18$total, na.rm = TRUE)
```
# create national percentage for 2018
```{r}
national_pct_2018 <- (total_el_2018/ total_cars_2018)* 100
```
# task m calculate the correlation coefficient
```{r}
cor(cardata$pct_expense, cardata$pct_elcar, use = "complete.obs")
```
# create he scatter plot
```{r}
plot(cardata$pct_expense, cardata$pct_elcar,main = "Relationship Between Expenses and Electric car %",xlab = "Percentage Expense(pct_expense)",ylab = "Percentage of Electric cars (pct_elcar)", col= "darkblue",pch= 19,las= 1)

```
# task n calculate the correlation coefficient for 2018
```{r}
cor_sent_el <- cor(cardata18$sent_index,cardata18$pct_elcar, use = "complete.obs")
```
# plot  the scatter plot
```{r}
plot(cardata18$sent_index, cardata18$pct_elcar,xlab = "centrality Index (sent_index)", ylab = "Pct Electric Cars (pct_elcar)", main = "Sent Index vs Electric Cars (2018) ",col= "darkblue", pch = 19, las= 1)
```
# task o calculate the correlation coefficient for 2018
```{r}
 plot(cardata18$sent_index, cardata18$pct_diesel,xlab = "centrality Index", ylab = "pct diesel", main = "Sent Index vs diesel Cars (2018)")
```

